---
title: "R Practice Problems - Solutions"
output: html_document
author: Zoe Vernon (many solutions provided by previous GSIs)
date: 08/29/2020
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Creating data structures 
### 1. Creating vectors 
#### a. 
```{r}
1:50 
seq(1, 50)
```

#### b. 
```{r}
!as.logical(1:50%%2)
```

#### c. 
```{r}
50:1
seq(50, 1)
```

#### d. 
```{r}
c(1:50, 49:1)
```

#### e. 
```{r}
-10:10
seq(-10, 10)
```
#### f. 
```{r}
3*1:16
seq(3, 48, by = 3)
```

#### g. 
```{r}
as.character(seq(3, 48, by = 3))
```

#### h. 
```{r}
rep(c("a", "b", "c", "d"), c(4, 3, 2, 1))
```

#### i. 
```{r}
as.factor(rep(c("a", "b", "c", "d"), c(4, 3, 2, 1)))
```

#### j. 
```{r}
seq(-1,1, length.out=200)
```

### 2. 
#### a. 
```{r}
x <- 1:50
```

#### b. 
```{r}
y <- cos(x)
```

#### c. 
```{r}
z <- tan(y)
```

#### d. 
```{r}
w <- y*z
```

#### e. 
```{r}
f <- x %in% 10:29
```

#### f. 
```{r}
df1 <- data.frame(x = x, y = y, z = z, w = w, f = f)
```

#### g. 
```{r}
names(df1) <- toupper(names(df1))
```

#### h. 
```{r}
# don't need to use x because rownames are alaready 1:50
df1 <- data.frame(y = y, z = z, w = w, f = f)

# can set rownames as well
df1 <- data.frame(y = y, z = z, w = w, f = f, row.names = x)
```

## Subsetting datastructures 
### 1. 
#### a. 
```{r}
m1 <- df1[, sapply(df1, is.numeric)]
```

#### b. 
```{r}
m2 <- m1[f, ]
```

#### c. 
```{r}
df2 <- df1[df1$z >= 0, names(df1) != 'z']
```

#### d. 
```{r}
df3 <- df1[-c(3, 17), ]
```

#### e. 
```{r}
df4 <- df1[seq(2, nrow(df1), 2), ]
```

## Vectorized calculations 
### 1. 
```{r}
x <- seq(1, 3, .1)
exp(2 * x) * x ^ sqrt(x)
```

### 2. 
#### a. 
```{r}
x <- matrix(0, nrow = 5, ncol = 5)
```

#### b. 
```{r}
row(x)
col(x)
```

#### c. 
```{r}
x[abs(row(x) - col(x)) == 1] = 1
```

#### d. 
```{r}
abs(row(x) - col(x))
```

### 3. 
#### a. 
```{r}
outer(0:4, 0:4, "+")
```

#### b. 
```{r}
outer(0:4, 0:4, "+") %% 5
```

## Using ```apply```, ```sapply```, and ```lapply```
### 1. 
#### a. 
```{r}
m1 <- matrix(data = runif(n = 30, min = 1, max = 100), nrow = 5, ncol = 6)
```

#### b. 
They are not the same, if we do it this way. R is column major
```{r}
m2 <- apply(X = m1, MARGIN = 1, FUN = function(x){ x / sum(x)})
dim(m1) == dim(m2) 
```

Now they match.
```{r}
m2 <- t(apply(X = m1, MARGIN = 1, FUN = function(x){x / sum(x)}))
dim(m1) == dim(m2)
```

#### c. 
```{r}
# check that the rows sum to 1
apply(X = m2, MARGIN = 1, FUN = function(x){sum(x) == 1})

# using rowSums
m2 <- m1 / rowSums(x = m1)
```

#### d. 
```{r}
m2 <- apply(X = m1, MARGIN = 2, FUN = function(x){x/sum(x)})

# dimensions are the same 
dim(m1) == dim(m2) 

# rows sum to 1
apply(X = m2, MARGIN = 2, FUN = function(x){sum(x) == 1})
```

### 2. 
#### a. 
```{r}
x <- seq.int(from = 1, to = 10, by = 0.1)
```

#### b. 
```{r}
y <- 2 * x + rnorm(n = length(x))
```

#### c. 
```{r}
plot(x, y)
```

#### d. 
```{r}
my_lm <- lm(y ~ x)
```

#### e. 
```{r}
lapply(X = my_lm, FUN = "class")
```

#### f. 
```{r}
sapply(X = my_lm, FUN = "class")
```

#### g. 
They are the same here, because a list is the best simplified type of object. ```sapply``` is a "simplified apply", ie, the output is simiplified to something deemed intelligent, like a vector or a list etc. 

## Functions 
### 1. 
#### 0. 
```{r}
absMedDiff <- function(x){
	return(sum(abs(x - median(x = x))))
}
```

### a. 
```{r}
absMedDiff2 <- function(x){
	# check if numeric
	stopifnot(is.numeric(x = x))
	
	# if numeric, do work and return
	return(sum(abs(x - median(x = x))))
}
```

#### b. 
```{r}
absMedDiff3a <- function(x, na.rm = FALSE){
	# check if numeric
	stopifnot(is.numeric(x = x))
	
	# if numeric, do work and return
	return(sum(abs(x - median(x = x, na.rm = na.rm)), na.rm = na.rm))
}
```

### 2. 
#### a. 
```{r}
x <- sample(x = 0:1, size = 100, replace = TRUE)
y <- sample(x = 0:1, size = 100, replace = TRUE)
```

#### b. 
It doesn't make sense to call set.seed within the function, because the function would always return the same sample.  
```{r}
sumHeads <- function(num_flips){
	# do flipping
	x <- sample(x = c(0,1), size = num_flips, replace = TRUE)
	
	# return heads
	return(sum(x))
}
```

#### c. 
```{r}
sums_vec <- replicate(n = 10000, expr = sumHeads(num_flips = 200))
```

#### d. 
```{r}
hist(sums_vec)
```

### 3. 
```{r}
basicFunc <- function(x, y, operation = "add"){
	switch(operation,
				 "add" = x + y,
				 "subtract" = x - y,
				 "multiply" = x * y,
				 "divide" = x / y,
				 "Warning: Not a designated option")
}
```

### 4. 
```{r}
cumSum <- function(x){
	# setup return vector
	ret_vec <- numeric(length = length(x) - 1)
	
	# set first element
	ret_vec[1] <- x[1]
	
	# loop over rest
	for(i in 2:length(x)){
		ret_vec[i] <- ret_vec[i - 1] + x[i]
	}
	
	# return result
	return(ret_vec)
}
```

## Loading (and saving) data 
### 1. 
Note, you will need to install the ```foreign``` package with ```install.packages("foreign")``` if it is not already.  The working directory is assumed to be the sections/01 folder in the stat243-fall-2020 Github. 
```{r}
library(foreign)
earnings <- read.dta("../data/heights.dta")
```

#### a. 
```{r}
class(earnings)
```

#### b. 
```{r}
str(earnings)
```

#### c. 
```{r}
length(earnings)
```

#### d. 
```{r}
dim(earnings)
```

#### e. 
```{r}
sapply(earnings, FUN = class)
```

#### f. 
```{r}
sapply(earnings[, grep("height", colnames(earnings))], summary)
```

#### g. 
```{r}

```

#### h. 
```{r}
hist(earnings$yearbn)
```

### 2. 
```{r}
# read csv
cpds <- read.csv("../data/cpds.csv")

# read text file 
state_income <- read.table("../data/stateIncome.txt", sep = "\t")
```

### 3. 
#### a. 
```{r}
ls()
```

#### b. 
There are number of formats to save data from R.  Below are a couple examples.
```{r}
# writing to a csv
write.csv(earnings, "../earnings.csv")

# writing to a text file 
write.table(m1, "../matrix.txt")

# write to an RData object, better for large data 
data_list <- list(earnings, m1, x, my_lm)
save(data_list, file = "../data_list.RData")
```

To load the .RData file you can use the ```load()``` function.  

